### 2019-09-26 aistats_paper/sections/experiments.tex
More WIP

We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
In total there were 2 million rows.
Where data were missing, we set the field to zero and the noise variance to a large value.
We set the projection $R_i$ to the identity matrix for every sample.
This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
We used a range of mixture component sizes $K$.
In practice we would want to select a value of $K$ by cross-validation.

@@

@@

>>>>

We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
Initial experiements were conducted using synthetic data sampled from a known mixture of Gaussians.

\subsection{Synthetic Data}

\todo{Description of synthetic setup}

\todo{Table of NLL results?}

\todo{Plot: Example 2D solution}

\todo{Plot: Obvious local maxima}

\todo{Plot: Denegenerate minibatch EM solution}

\subsection{Gaia Catalogue}

To evaluate each method with a real application, we used a random sample of rows from the Gaia DR2 source table~\citep{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
In total there were 2 million rows.
Where data were missing, we set the field to zero and the noise variance to a large value.
We set the projection $R_i$ to the identity matrix for every sample.
This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.

@@

\subsubsection{Variable K}

@@

\subsubsection{Variable N}

### 2019-09-26 aistats_paper/sections/methods.tex
More WIP

@@

>>>>

Here we present an alternative update for $V_j$ that is less prone to numerical instability, and show that it is equivalent to Equation~\ref{eqn:mstep}.
For clarity we drop the component indicator $j$ from the parameters, and add indicators $t$ and $t-1$ to distinguish between current and previous estimates of parameters.

First, we define an adjustment operation,
\begin{align}
  \begin{split}
  \text{adjust}(V, s, \bc, \bd) =& \ sV + \frac{1}{2}(\bc - \bd)(\bc + \bd)^T \\
  &+ \frac{1}{2}(\bc + \bd)(\bc - \bd)^T  \label{eq:recentre1}
  \end{split} \\
  =& \ s(V + \bc\bc^T) - \bd\bd^T,\label{eq:recentre2}
\end{align}
which can be thought of as recentering a scaled variance around a new mean.
Equation~\ref{eq:recentre1} is how we actually compute the adjustment, to minimise taking small differences between large values, whilst Equation~\ref{eq:recentre2} shows the identity we are interested in.

In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
We also compute minibatch-specific parameters using exact sums over the minibatch:
\begin{align}
q_b &= \sum_i^M r_i, \\
\bm_b &= \frac{\sum_i^M r_i \bx_i}{q_b},\\
V_b &= \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
\end{align}
We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
\begin{align}
\begin{split}
V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) \\
&+ \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}
\end{split} \\
&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
\end{align}
Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.

@@

\subsection{Stochastic Split-Merge}

\todo{Add description of proposed stochastic split-merge}
